Optimizing Abstract Abstract Machines

### 2012-11-13 paper/paper.tex
Less detail in intro, more in evaluation. Better table format.

\section{At a Glance}

In the following sections, we develop an analyzer for a core
functional language and follow a series of steps of abstraction and
optimization.  At each step we evaluate the implementation on an
example that computes with Church numerals, shown in
figure~\ref{fig:church}.
%% \footnote{It is
%%   written here using Scheme notation and makes use of the shorthands:
%% %
%% \begin{alltt}
%%    (define (((f x) y) z) ...) \(\equiv\)
%%
%%    (define f (\(\lambda\) (x) (\(\lambda\) (y) (\(\lambda\) (z) ...))))
%% \end{alltt}
%%
%% \begin{alltt}
%%    \church{n} \(\equiv\) (\(\lambda\) (s) (\(\lambda\) (z) (s\(\subscript1\) (s\(\subscript2\) ... (s\(\subscript{n}\) z)))))
%% \end{alltt}}
The example, while small, is informative:
\begin{enumerate}
\item it can be written in most modern programming languages, and
%
\item it stresses an analyzer's ability to deal with complicated
environment and control structure arising from the use of higher-order
functions to model arithmetic.
\end{enumerate}

We start, in section~\ref{sec:aam}, by developing an abstract
interpreter according to the abstracting abstract machines approach
and observe the resulting implementation cannot analyze the example in
under 30 minutes.  In section~\ref{sec:baseline}, we perform a further
abstraction by widening the store.  The resulting analyzer sacrifices
precision for speed and is able to analyze the example in about 1
minute.  Without this abstraction step the analysis is exponential due
to per-state store variance.  This step is described by Van Horn and
Might~\cite[\S 3.5--6]{dvanhorn:VanHorn2012Systematic} and is
necessary to make even small examples feasible.  We therefore take the
widened interpreter as the baseline for our evaluation.

In section~\ref{sec:opt} we give a series of simple abstractions and
implementation techniques that, in total, drop the analysis time to
only 138 milliseconds, a speedup of 475 times.
Figure~\ref{fig:churchtime} shows the step-wise improvement of the
analysis time for this example.

@@

In section~\ref{sec:impl}, we describe a more complete implementation
scaled up and capable of analyzing Scheme benchmark programs.

In section~\ref{sec:eval} we evaluate this implementation against a
set of benchmark programs drawn from the literature.  Many benchmarks
cause the baseline analyzer to take longer than 30 minutes or to
consume more 1 gigabit of memory, at which point the analysis is
stopped.  This is the case for the largest benchmark program, which is
3,500 lines of code and takes under a minute in the most optimized
analyzer.  For those benchmarks that did complete on the baseline, the
optimized analyzer outperformed the baseline by a factor of
% 475
two to
% 4,382
three orders of magnitude.

@@

\begin{figure}
\begin{tiny}
\begin{alltt}
Program  |  LOC | B. Time | O. Time | B. Space | O. Space | B. Rate | O. Rate
----------------------------------------------------------------------------
nucleic  | 3508 |
matrix   |  747 |
nbody    | 1434 |
earley   |  667 |
maze     |  681 |
church   |   41 |
lattice  |  214 |
boyer    |  642 |
mbrotz   |   68 |
\end{alltt}
\end{tiny}
\end{figure}

\begin{itemize}

@@

\end{itemize}

@@

@@

The analysis most similar to that presented in section~\ref{sec:aam}
is the infinitary control-flow analysis of Nielson and
Nielson~\cite{dvanhorn:nielson-nielson-popl97} and the unified
treatment of flow analysis by Jagannathan and
Weeks~\cite{dvanhorn:jagannathan-weeks-popl95}.  Both are
parameterized in such a way that in the limit, the analysis is
equivalent to an interreter for the language, just as is the case
here.  What is different is that both give a constraint-based
formulation of the abstract semantics rather than a finite machine
model.

@@

%% \acks Sam Tobin-Hochstadt for encouragement and feedback -- he was
%% the first to prompt me to look into how make effective
%% implementations of the AAM approach.

>>>>

\section{At a glance}

In the following sections, we develop an analyzer for a core
functional language and follow a series of steps of abstraction and
optimization.  At each step we evaluate the implementation on an
example that computes with Church numerals, shown in
figure~\ref{fig:church}.
%% \footnote{It is
%%   written here using Scheme notation and makes use of the shorthands:
%% %
%% \begin{alltt}
%%    (define (((f x) y) z) ...) \(\equiv\)
%%
%%    (define f (\(\lambda\) (x) (\(\lambda\) (y) (\(\lambda\) (z) ...))))
%% \end{alltt}
%%
%% \begin{alltt}
%%    \church{n} \(\equiv\) (\(\lambda\) (s) (\(\lambda\) (z) (s\(\subscript1\) (s\(\subscript2\) ... (s\(\subscript{n}\) z)))))
%% \end{alltt}}
The example, while small, is informative:
\begin{enumerate}
\item it can be written in most modern programming languages, and
%
\item it stresses an analyzer's ability to deal with complicated
environment and control structure arising from the use of higher-order
functions to encode arithmetic.
\end{enumerate}

We start, in section~\ref{sec:aam}, by developing an abstract
interpreter according to the AAM approach.  Without further
abstraction, the analysis is exponential due to per-state store
variance and thus cannot analyze the example in a reasonable amount of
time.  In section~\ref{sec:baseline}, we perform a further abstraction
by widening the store.  The resulting analyzer sacrifices precision
for speed and is able to analyze the example in about 1 minute.  This
step is described by Van Horn and Might~\cite[\S
  3.5--6]{dvanhorn:VanHorn2012Systematic} and is necessary to make
even small examples feasible.  We therefore take the widened
interpreter as the baseline for our evaluation.

Section~\ref{sec:opt} gives a series of simple abstractions and
implementation techniques that, in total, speed up the analysis by
nearly a factor of 500, dropping the analys time to a fraction of a
second.  Figure~\ref{fig:churchtime} shows the step-wise improvement
of the analysis time for this example.

@@

Section~\ref{sec:eval} describes the evaluation of each optimiztion
technique applied to an implementation supporting a more realistic set
of features, including mutation, first-class control, compound data, a
full numeric tower and many more forms of primitive data and
operations.
%
We evaluate this implementation against a set of benchmark programs
drawn from the literature.
%
For all benchmarks, the optimized analyzer outperforms the baseline
by at least a factor of
% 475
two to
% 4,382
three orders of magnitude.

@@

The benchmarks make use of mutation, first-class control (in the form
of Scheme's {\tt call/cc} operator), compound data such as lists and
vectors, many more primitive forms of data and operations such as
Scheme's numeric tower including exact rationals, complex, and
floating point numbers.

Many benchmarks cause the baseline analyzer to take longer than 30
minutes or to consume more 1 gigabit of memory, at which point the
analysis is stopped.  This is the case for the largest benchmark
program, which is 3,500 lines of code and takes under a minute in the
most optimized analyzer.  For those benchmarks that did complete on
the baseline, the optimized analyzer outperformed the baseline by a
factor of 475 to 4,382.

\begin{figure}
\begin{tabular}{@{}l|r|r|r|r|r|r|r@{}}
Program & LOC
& \multicolumn{2}{c|}{Time (ms)}
& \multicolumn{2}{c|}{Space (mb)}
& \multicolumn{2}{c@{}}{Speed (state/sec)}
\\
\hline\hline
nucleic &  3508 & 0 & 0 & 0 & 0 & 0 & 0\\
matrix  &  747  & ? & ? & ? & ? & ? & ?\\
nbody   & 1434  & ? & ? & ? & ? & ? & ?\\
earley  &  667  & ? & ? & ? & ? & ? & ?\\
maze    &  681  & ? & ? & ? & ? & ? & ?\\
church  &   41  & ? & ? & ? & ? & ? & ?\\
lattice &  214  & ? & ? & ? & ? & ? & ?\\
boyer   &  642  & ? & ? & ? & ? & ? & ?\\
mbrotz  &   68  & ? & ? & ? & ? & ? & ?
\end{tabular}
\caption{Overview performance comparison between baseline and
  optimized analyzer.}
\label{fig:bench-overview}
\end{figure}

\begin{enumerate}  %% Maybe use ``dictionary'' style enumeration.

@@

\end{enumerate}

@@

\paragraph{Comparison with other flow analysis implementations}

[Computes results more like Earl, et al.'s 0CFA, which times out on
  the Church numeral benchmark and so even though it offers a fair
  point of comparison, a more thorough evaluation is likely to
  uninformative as the other benchmarks are highly likely to timeout
  as well (and it would require significant effort to extend their
  implementation with the features needed to analyze our benchmark
  suite).  That implementation is evaluated against much smaller
  benchmarks: the largest program consists of 63 \emph{subexpressions}
  and times out after 30 minutes.]

Vardoulakis and Shivers evaluate their CFA2
analyzer~\cite{dvanhorn:Vardoulakis2011CFA2} against a variant of 0CFA
defined in their framework.  [NEED TIMING FOR 0CFA ON CHURCH.]  The
Church example is the largest benchmark Vardoulakis and Shivers
consider.  More work would be required to scale the analyzer to the
set of features required by our benchmarks.

[Should probably talk about mCFA too.]

The only analyzers we were able to find that proved capable of
analyzing the full suite of benchmarks considered here were the Soft
Typing system of Wright [CITE] and, in many ways its successor, the
Polymorphic splitting system of Wright and
Jagannathan~\cite{dvanhorn:wright-jagannathan-toplas98}.\footnote{This
  is not a coincidence; these papers set a high standard for
  evaluation, which we consciously aimed to approach.  Many of our
  benchmarks are drawn from these papers.}  Unfortunately, these
analyses compute an inherently different and uncomparable form of
analysis.  Consequently, we have omitted a complete comparison with
these implementations,.  The AAM approach provides more precision in
terms of temporal-ordering of program states, which comes at a cost
that can be avoided in constraint-based approaches.  Consequently
implementation techniques cannot be ``ported'' between these two
approaches.  However, our optimized implementation is within an order
of magnitude of the performance of Wright and Jaganathan's analyzer.
Although we would like to improve this to be more competitive, the
optimized AAM approach still has many strengths to recommend it in
terms of precision, ease of implementation and verification, and rapid
design.

@@

Besides the original AAM work, the analysis most similar to that
presented in section~\ref{sec:aam} is the infinitary control-flow
analysis of Nielson and Nielson~\cite{dvanhorn:nielson-nielson-popl97}
and the unified treatment of flow analysis by Jagannathan and
Weeks~\cite{dvanhorn:jagannathan-weeks-popl95}.  Both are
parameterized in such a way that in the limit, the analysis is
equivalent to an interreter for the language, just as is the case
here.  What is different is that both give a constraint-based
formulation of the abstract semantics rather than a finite machine
model.

@@

%% \acks Sam Tobin-Hochstadt for encouragement and feedback -- he was
%% the first to prompt us to look into how make effective
%% implementations of the AAM approach.

%% Vincent St Amour for feedback on early drafts.
%% Greg Morrisset and Matthias Felleisen for discussions.
